Model Checking Differentially Private Properties

### 2018-01-27 dpCTL.tex
structures and some examples

The logic $\dpCTL$ is designed to specify differentially private
mechanisms. We introduce the differentially private path quantifer
$\dpriv{\epsilon}{\delta}$ in $\dpCTL$. For any path formula
$\phi$, a state $s$ in a Markov chain $K$ satisfies
$\dpriv{\epsilon}{\delta} \phi$ if the probability of having $\phi$-paths 
from $s$ is close to the probabilities of 
having $\phi$-paths from its neighboring states. 
We give semantics for both Markov chains and Markov decision processes.

>>>>

The logic $\dpCTLstar$ is designed to specify differentially private
mechanisms. We introduce the differentially private path quantifer
$\dpriv{\epsilon}{\delta}$ in $\dpCTLstar$. For any path formula
$\phi$, a state $s$ in a Markov chain $K$ satisfies
$\dpriv{\epsilon}{\delta} \phi$ if the probability of having $\phi$-paths
from $s$ is close to the probabilities of
having $\phi$-paths from its neighboring states.
%We give semantics for both Markov chains and Markov decision processes.

### 2018-01-27 examples.tex
first try
added some properties
structures and some examples

@@

A simple way to design differentially private mechanisms is to add 
random noises. Consider the survey question: have you been diagnosed
with the disease $X$? In order to protect surveyees' privacy, each
answers the question as follows. The surveyee first flips a
coin. If it is tail, she answers the question truthfully. Otherwise, 
she randomly answers \textit{Yes} or \textit{No}
uniformly~\cite{DR:14:AFDP}. Figure~\ref{figure:2-dp} shows the survey
mechanism and its corresponding Markov chain. In the figure, the states
$+$ and $-$ denote the diagnoses are positive or negative
respectively; the states $Y$ and $N$ denote answers to the survey
question. The random noise boosts the probability of answering
\textit{Yes} or \textit{No} to at least $1/4$ regardless of
diagnoses. Inferences on individual diagnosis can be plausibly denied.

Let us analyze the survey mechanism briefly. The data universe
$\mathcal{X}$ is $\{ +, - \}$. The mechanism $M$ is a randomized
algorithm with inputs in $\mathcal{X}$ and outputs in $\{ Y, N
\}$. For any $x \in \mathcal{X}$, we have $1/4 \leq \Pr[M(x) = Y] \leq
3/4$. Hence for any neighboring $x, x' \in \mathcal{X}$, $\Pr[M(x) =
Y] \leq 3/4 = 3 \cdot 1/4 \leq 3 \Pr[M (x') =
Y]$. Similarly, $\Pr[M (x) = N] \leq 3 \Pr[M (x') = N]$ for any
neighboring $x, x' \in \mathcal{X}$. The survey mechanism is
$(3, 0)$-differentially private.

More sophisticated differentially private mechanisms are
available. Consider a query 
$f : \mathcal{X}^n \rightarrow \bbfZ^{\geq 0}$. Let $\alpha \in (0, 1)$. 
The \emph{$\alpha$-geometric mechanism} 
outputs $f(x) + Y$ where $Y$ is a random variable with the geometric
distribution~\cite{GRS:09:UUPM,GRS:12:UUPM} :
\[
\Pr[Y = y] = \frac{1 - \alpha}{1 + \alpha}\alpha^{|y|}
\textmd{ for } y \in \bbfZ.
\]
The $\alpha$-geometric mechanism is oblivious since it has the same
output distribution on any inputs $x, x'$ with $f (x) = f
(x')$. Moreover, it is also $(\alpha^{\Delta (f)}, 0)$-differentially
private for any query $f$ with sensitivity $\Delta (f)$. The range of
the mechanism 
however is $\bbfZ$. It may give nonsensical outputs such as
negative integers for queries with positive ranges.

@@

\begin{figure}
  \centering
  \begin{subfigure}{.40\columnwidth}
    \resizebox{\columnwidth}{!}{
    \begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node
      distance=2cm,node/.style={circle,draw}]
      \node[node] (p) at ( 0,  1.5) { $+$ };
      \node[node] (q) at ( 0, -1.5) { $-$ };
      \node[node] (F) at (-1.5,  0) { $Y$ };
      \node[node] (T) at ( 1.5,  0) { $N$ };

      \path
      (p) edge [bend right=45] node [left] { $L, .75$ } (F)
      (p) edge [bend left=45] node [right] { $L, .25$ } (T)
      (p) edge [bend left=45] node [right=-12,above] { $H, .8$ } (F)
      (p) edge [bend right=45] node [right=12,above] { $H, .2$ } (T)

      (q) edge [bend left=45] node [left] { $L, .25$ } (F)
      (q) edge [bend right=45] node [right] { $L, .75$ } (T)
      (q) edge [bend right=45] node [left=12,below] { $H, .2$ } (F)
      (q) edge [bend left=45] node [right=12,below] { $H, .8$ } (T)

      (F) edge [loop left] node [below] { $-, 1$ } (F)
      (T) edge [loop right] node [below] { $-, 1$ } (T)
      ;
      \end{tikzpicture}
    }
    \caption{Markov Decision Process}
    \label{figure:simple-mdp-mdp}
  \end{subfigure}
  \begin{subfigure}{.58\columnwidth}
    \resizebox{\columnwidth}{!}{
    \begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node
      distance=2cm,node/.style={circle,draw,inner sep=1pt}]
      \node[node] (pq) at (-2.1, 2.1) { $+-$ };
      \node[node] (qp) at ( 2.1,-2.1) { $-+$ };
      \node[node] (RR) at (-2.1,-2.1) { $YY$ };
      \node[node] (BB) at ( 2.1, 2.1) { $NN$ };
      \node[node] (RB) at (- .7,- .7) { $YN$ };
      \node[node] (BR) at (  .7,  .7) { $NY$ };

      \path
      (pq) edge [bend right=5] node [left=1,below=12,rotate=270] { $L, .1875$ } (RR)
      (pq) edge [bend left=5]  node [above=5,right=1] { $L, .1875$ } (BB)

      (pq) edge [bend left=5]  node [right=4,rotate=270] { $H, .16$ } (RR)
      (pq) edge [bend right=5] node [below=5,right=1] { $H, .16$ } (BB)

      (pq) edge [bend left=5]  node [left,below=10,rotate=295]  { $L, .5625$ } (RB)
      (pq) edge [bend right=5] node [left,below,rotate=330]  { $L, .0625$ } (BR)

      (pq) edge [bend right=5] node [right=2,above,rotate=300] { $H, .64$ } (RB)
      (pq) edge [bend left=5]  node [right,above,rotate=340] { $H, .04$ } (BR)

      (qp) edge [bend right=5] node [above=5,left=1] { $L, .1875$ } (RR)
      (qp) edge [bend left=5]  node [above=30,left=5,rotate=90] { $L, .1875$ } (BB)

      (qp) edge [bend left=5]  node [below=5,left=1] { $H, .16$ } (RR)
      (qp) edge [bend right=5] node [above,right=5,rotate=90] { $H, .16$ } (BB)

      (qp) edge [bend left=5]  node [left,below,rotate=330]  { $L, .0625$ } (RB)
      (qp) edge [bend right=5] node [left=3,below=3,rotate=295]  { $L, .5625$ } (BR)

      (qp) edge [bend right=5] node [left=5,above,rotate=330] { $H, .04$ } (RB)
      (qp) edge [bend left=5]  node [right=1,above=1,rotate=295] { $H, .64$ } (BR)

      (RR) edge [loop left] node [above] { $-, 1$ } (RR)
      (BB) edge [loop right] node [above] { $-, 1$ } (BB)
      (RB) edge [loop below] node [left] { $-, 1$ } (RB)
      (BR) edge [loop above] node [right] { $-, 1$ } (BR)
      ;
      \end{tikzpicture}
\hide{
    \begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node
      distance=2cm,node/.style={circle,draw,inner sep=1pt}]
      \node[node] (pq) at (   0,   0) { $+-$ };
      \node[node] (RR) at (-  2,   0) { $YY$ };
      \node[node] (BB) at (   2,   0) { $NN$ };
      \node[node] (RB) at (   0, 1.5) { $YN$ };
      \node[node] (BR) at (   0,-1.5) { $NY$ };

      \path
      (pq) edge [bend right=15] node [above] { $L, .1875$ } (RR)
      (pq) edge [bend left=15]  node [above] { $L, .1875$ } (BB)

      (pq) edge [bend left=15]  node [below] { $H, .16$ } (RR)
      (pq) edge [bend right=15] node [below] { $H, .16$ } (BB)

      (pq) edge [bend left=15]  node [left]  { $L, .5625$ } (RB)
      (pq) edge [bend right=15] node [left]  { $L, .0625$ } (BR)

      (pq) edge [bend right=15] node [right] { $H, .64$ } (RB)
      (pq) edge [bend left=15]  node [right] { $H, .04$ } (BR)

      (RR) edge [loop below] node [below] { $-, 1$ } (RR)
      (BB) edge [loop above] node [above] { $-, 1$ } (BB)
      (RB) edge [loop left] node [left] { $-, 1$ } (RB)
      (BR) edge [loop right] node [right] { $-, 1$ } (BR)
      ;
      \end{tikzpicture}
}
    }
    \caption{Product Markov Decision Process}
    \label{figure:simple-mdp-product}
  \end{subfigure}  
  \caption{Markov Decision Process and its Self-Product}
  \label{figure:simple-mdp}
\end{figure}

Mechanisms are not necessarily closed randomized algorithms; they may
perform different compution on users' requests. 
We use Markov decision processes to model such 
interactive mechanisms. Specifically, external inputs are modeled by
actions. Behaviors associated with different inputs are modeled by
distributions associated with actions.

Consider again the survey mechanism. Suppose we would like to design
an interactive mechanism which adjusts random noises on surveyers'
requests. When the surveyer requests low-accuracy answers, the
surveyee uses the survey mechanism as before. When high-accuracy
answers are requested, the surveyee answers \textit{Yes} with
probability $4/5$ and \textit{No} with probability $1/5$ when she has
positive diagnosis. She answers \textit{Yes} with probability $1/5$
and \textit{No} with probability $4/5$ when she is not
diagnosed with the disease X. This gives an interactive mechanism
corresponding to the Markov decision process shown in
Figure~\ref{figure:simple-mdp-mdp}.

In the figure, the states $+$,
$-$, $Y$, and $N$ are interpreted as before. The actions $L$ and $H$
denote low- and high-accuracy requests respectively. Let $M_H$ denote
the Markov chain derived by high-accuracy requests.
Observe that $1/5 \leq \Pr[M_H(x) = Y] \leq 4/5$ for any $x
\in \mathcal{X}$. Hence $\Pr[M_H (x) = Y] \leq 4/5 = 4 \cdot 1/5 \leq
4 \Pr[M_H (x') = Y]$ for any neighboring $x, x' \in
\mathcal{X}$. Similarly, $\Pr[M_H (x) = N] \leq 4 \Pr[M_H (x') =
N]$. The high-accuracy mechanism is $(4,0)$-differentially private. 
The privacy guarantees vary from accuracy requests.

>>>>

As discussed in previous section, a simple way to design differentially private mechanisms is to add
random noises. 
Thus, in this section, we discuss how to model differentially private mechanisms using Markov chains. We explain the main idea first with some examples.

\paragraph{Survey Mechanisms.}

@@

Consider the survey question: have you been diagnosed
with the disease $X$? In order to protect surveyees' privacy, each
answers the question as follows. The surveyee first flips a
coin. If it is tail, she answers the question truthfully. Otherwise,
she randomly answers \textit{Yes} or \textit{No}
uniformly~\cite{DR:14:AFDP}.

\begin{example}\label{exa:survey}
  Figure~\ref{figure:2-dp} shows the survey
mechanism and its corresponding Markov chain. In the figure, the states
$+$ and $-$ denote the diagnoses are positive or negative
respectively; the states $Y$ and $N$ denote answers to the survey
question. We have $AP=\{Y,N\}$, $L(Y)=\{Y\}$, $L(N)=\{N\}$, and $L(+)=L(-)=\emptyset$. \lz{draw it on the picture?}

The random noise boosts the probability of answering
\textit{Yes} or \textit{No} to at least $1/4$ regardless of
diagnoses. Inferences on individual diagnosis can be plausibly denied.
\end{example}

Let us recall briefly the analysis of the survey mechanism. The data universe
$\mathcal{X}$ is $\{ +, - \}$. The mechanism $M$ is a randomized
algorithm with inputs in $\mathcal{X}$ and outputs in $\{ Y, N
\}$. For any $x \in \mathcal{X}$, we have $1/4 \leq \Pr[M(x) = Y] \leq
3/4$. Hence for any neighboring $x, x' \in \mathcal{X}$, $\Pr[M(x) =
Y] \leq 3/4 = 3 \cdot 1/4 \leq 3 \Pr[M (x') =
Y]$. Similarly, $\Pr[M (x) = N] \leq 3 \Pr[M (x') = N]$ for any
neighboring $x, x' \in \mathcal{X}$. The survey mechanism is
$(3, 0)$-differentially private.

\paragraph{Truncated $\alpha$-Geometric Mechanism.}
More sophisticated differentially private mechanisms are
available. Consider a query
$f : \mathcal{X}^n \rightarrow \bbfZ^{\geq 0}$. Let $\alpha \in (0, 1)$.
The \emph{$\alpha$-geometric mechanism}
outputs $f(x) + Y$ where $Y$ is a random variable with the geometric
distribution~\cite{GRS:09:UUPM,GRS:12:UUPM} :
\[
\Pr[Y = y] = \frac{1 - \alpha}{1 + \alpha}\alpha^{|y|}
\textmd{ for } y \in \bbfZ.
\]
The $\alpha$-geometric mechanism is oblivious since it has the same
output distribution on any inputs $x, x'$ with $f (x) = f
(x')$. Moreover, it is also $(\alpha^{\Delta (f)}, 0)$-differentially
private for any query $f$ with sensitivity $\Delta (f)$. The range of
the mechanism
however is $\bbfZ$. It may give nonsensical outputs such as
negative integers for queries with positive ranges.

@@

\subsection{Markov Chains for Differentially Private Mechanisms}
Now we describe our main ideas for the modelling differentially private mechanisms. 
A state $s$ stores at least the a database $x\in \mathcal{X}^n$, which is the data hold by a trusted \emph{curator}. A transition corresponds to the data analysis mechanisms.
States are typically labelled with the corresponding outputs
the curator generates.

In the differentially privacy context, we differentiate between offline and online models. 
In the offline, or non-interactive, model the curator releases the outputs only once and plays no further role. Thus the models we constructed above are offline models. On the contrary, in the online, or interactive, model the curator can permit further queries. In this setting it is obvious that queries do not change database themselves. To maintain differentially privacy, the curator however can decide to change data analysis mechanisms accordingly. For instance, after a certain output is disclosed or a certain number of queries, it may decide to disable further queries.

Below we describe an online model.

\begin{figure}
    \centering
    \begin{subfigure}{.4\columnwidth}
\resizebox{\columnwidth}{!}{
    \begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node
    distance=1cm,node/.style={circle,draw}]
    \node[node] (p) at ( -1.5,  0) { $+$ };
    \node[node] (Y1) at ( 0, .8) { $Y_1$ };
    \node[node] (N1) at ( 0, -.8) { $N_1$ };
    \node[node] (Y2) at (1.5,  .8) { $Y_2$ };
    \node[node] (N2) at ( 1.5, -.8) { $N_2$ };

    \path
    (p) edge node [above] { $\frac{3}{4}$ } (Y1)
    (p) edge node [below] { $\frac{1}{4}$ } (N1)
    (Y1) edge node [above] { $\frac{3}{4}$ } (Y2)
    (Y1) edge node [above] { $\frac{1}{4}$ } (N2)
    (N1) edge node [below] { $\frac{3}{4}$ } (Y2)
    (N1) edge node [below] { $\frac{1}{4}$ } (N2)
    ;
    \end{tikzpicture}
}
%        \caption{Markov Chain of }
%        \label{figure:2-dp-mc1}
    \end{subfigure}
    \hspace{.05\columnwidth}
    \begin{subfigure}{.4\columnwidth}
        \resizebox{\columnwidth}{!}{
            \begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node
    distance=1cm,node/.style={circle,draw}]
    \node[node] (n) at ( -1.5,  0) { $-$ };
    \node[node] (Y1) at ( 0, .8) { $Y_1$ };
    \node[node] (N1) at ( 0, -.8) { $N_1$ };
    \node[node] (Y2) at (1.5,  .8) { $Y_2$ };
    \node[node] (N2) at ( 1.5, -.8) { $N_2$ };

    \path
    (n) edge node [above] { $\frac{1}{4}$ } (Y1)
    (n) edge node [below] { $\frac{3}{4}$ } (N1)
    (Y1) edge node [above] { $\frac{1}{4}$ } (Y2)
    (Y1) edge node [above] { $\frac{3}{4}$ } (N2)
    (N1) edge node [below] { $\frac{1}{4}$ } (Y2)
    (N1) edge node [below] { $\frac{3}{4}$ } (N2)
    ;
            \end{tikzpicture}
        }
%        \caption{Markov Chain of}
%        \label{figure:2-dp-mc2}
    \end{subfigure}
    \caption{Markov Chain of }
    \label{figure:2-dp-mc}
\end{figure}
\begin{example}
  First, assume that the curator allows at most two survey queries. The corresponding model is depicted in Figure~\ref{figure:2-dp-mc}, where states $Y_1,Y_2$ are labelled by $Y$, and $N_1,N_2$ are labelled by $N$.
\end{example}

\lz{tbd, the complex threshold example we discussed, by Depeng}

### 2018-01-27 semantics.tex
structures and some examples

Other than the differentially private operator, the semantics of
$\dpCTL$ is standard~\cite{BK:08:PMC}.
To intuit the semantics of $\dpriv{\epsilon}{\delta} \phi$,
recall that  $\myPr{s}{K}{N}{\phi}$ is the probability of having
paths satisfying $\phi$ from $s$. Thus, a state $s$ satisfies
$\dpriv{\epsilon}{\delta} \phi$ if having paths satisfying $\phi$
from $s$ is $(\epsilon, \delta)$-close to having paths satisfying
$\phi$ from every neighbor of $s$. In other words, it is
probabilistically similar to observe paths satisfying $\phi$ from $s$
and from all neighbors of $s$.

\section{Model Checking Algorithms}
The model checking algorithms for $\dpCTL$ follow the classical
algorithms for $\PCTL$ by computing the states satisfying
sub state-formulae inductively~\cite{BK:08:PMC}. It hence suffices to
consider the inductive step where the states satisfying the subformula
$\dpriv{\epsilon}{\delta} (\phi)$ is computed.

@@

\section{Specifying Properties in $\dpCTL$}
In this section we describe what type of properties in the differential privacy context can be expressed using $\dpCTL$.

\paragraph{Standard Differential Privacy.} 
The formula $\dpriv{3}{0} (\X Y)$ holds in state $s$ if for  $t$ with $sNt$, the probability of satisfying $\X Y$ from $s$ and $t$ satisfy:
$\myPr{s}{K}{N}{\X Y}\le 3\myPr{t}{K}{N}{\X Y}$. The formula $\dpriv{3}{0} (\X Y)\wedge \dpriv{3}{0} (\X N)$ thus specifies the differential privacy property for states $+$ and $-$ in Example~\ref{exa:survey}.

Further, consider the Markov chain in
Figure~\ref{figure:geometric-mechanism-markov-chain}. Assume
$L(out_i) = \{ out_i \}$ and $\Delta (f) = 1$. Then $in_k$ and $in_l$ are
neighbors iff $| k - l | \leq 1$\lz{put above sentence to the def of the model}. Define the $\dpCTL$ formula
$\psi = \dpriv{.5}{0} (\F out_0) \wedge \dpriv{.5}{0} (\F out_1) \wedge
\cdots \wedge \dpriv{.5}{0} (\F out_5)$\lz{I think we better use $\X$ operator here ;)}. If the state $in_k$ satisfies
$\psi$ for $k = 0, \ldots, 5$, then the $.5$-geometric mechanism is
$(0.5, 0)$-differentially private.

\paragraph{Compositional Properties.}
Compositional aspect is one of the building block for differentially private algorithms. According to the compositional theorem \cite{}, for $(\epsilon_1,0)$-differentially private algorithm $M_1$ and
$(\epsilon_2,0)$-differentially private algorithm $M_2$, their combination $M$ defined by $M(x)=(M_1(x), M_2(x))$ is $(\epsilon_1\epsilon_2,0)$-differentially private. The degradation is rooted in the repeated computation. To illustrate this property, we consider the extended example below \lz{here example of the survey example, expanded twice: we should discuss the graph before it is drawn...}

We consider the formula $\dpriv{9}{0} (\X (Y\wedge \X Y))$. A path satisfies $\X (Y\wedge \X Y)$ if the second state satisfies $Y$ and the third state satisfies $Y$ as well. We verify that this formula is satisfied  for state $+$ and $-$. Moreover, the bound $\epsilon=9$ is bound, since \lz{tbd once graph is given}. Finally, the formula $\wedge_{a_1,a_2}\dpriv{9}{0} (\X (a_1\wedge \X a_2))$ corresponds to the desired compositional property for the model, where $a_1,a_2$ range over all atomic propositions $\{Y,N\}$.

We consider another simpler formula for comparision: $\dpriv{3}{0} (\X \X Y)$. In this case we donot have privacy lost, even though we are querying twice: the reason is that the output of the first query is not utilized at all. It is easy to verify that it is indeed satisfied by $+$ and $-$.

\paragraph{Advanced Compositional Properties}
One of the advantage of applying model checking is that we may get tighter bounds for compositional properties. Consider our survey example, and the property $\dpriv{1}{.5} (\X Y)$. Obviously, it holds in states $+$ and $-$. A careful check ensures that one cannot decrease the value of $\epsilon_1=1$ or $\delta_1=.5$ without increasing the other one.

Now we consider the formula $\dpriv{\epsilon_2}{\delta_2} (\X (Y\wedge \X Y))$ under the model in ???. Applying the compositional theorem\lz{Bow-Yaw, as we only refer to the book, I wonder whether using the notion $e^\epsilon$ is easier to explain?}, one can choose $\epsilon_2=\epsilon_1^2=1$ and $\delta_2=2\delta_1=1$. However, we can check easily that one can get better privacy parameter $(1,0.5)$ or $(0,\frac{9}{16})$.

\paragraph{Infinite ...}
Note that $\dpCTL$ generalizes differential privacy to infinite
behaviors naturally. Classical differential privacy discusses properties about
randomized terminating mechanisms. Sophisticated mechanisms for
streaming data have also been discussed in social networks and
internet of things. Such mechanisms inherently have infinite
computation. Their privacy guarantees are approximated by those of the
prefixes of computation.
With \dpCTL, the subformula $\phi$ in $\dpriv{\epsilon}{\delta} \phi$
can be an arbitrary $\dpCTL$ path formula. One can specify
$(\epsilon, \delta)$-closedness over infinite behaviors naturally. For instance,
$\dpriv{\epsilon}{\delta} (\G \PJ{[1,1]} (\F \mathit{high}))$ \lz{now we can use $G F high$}
specifies states having $(\epsilon, \delta)$-close
infinitely often $\mathit{high}$ behaviors to its neighbors.

\section{Markov Decision Processes}
The semantics over Markov chains generalizes to Markov decision
processes as well. Note that two states with different enabled actions
are trivially distinguishable; no privacy can be preserved. Without
loss of generality, we consider reactive Markov decision processes.
Let $M = (S, \Act, \wp, L)$ be a reactive Markov
decision process. Define the satisfaction
relation $M, \neighbor{S}, s \models \Phi$ as follows.
\begin{eqnarray*}
  M, \neighbor{S}, s \models \top\\
  M, \neighbor{S}, s \models p
  & \textmd{ if } &
  p \in L(s)\\
  M, \neighbor{S}, s \models \neg \Phi
  & \textmd{ if } &
  M, \neighbor{S}, s \not\models \Phi\\
  M, \neighbor{S}, s \models \Phi_0 \wedge \Phi_1
  & \textmd{ if } &
  M, \neighbor{S}, s \models \Phi_0 \textmd{ and }
  M, \neighbor{S}, s \models \Phi_1\\
  M, \neighbor{S}, s \models \PJ{J} \phi
  & \textmd{ if } &
  \myPr{s}{M_{\scheduler{S}}}{\neighbor{S}}{\phi} \in J
  \textmd{ for every scheduler } \scheduler{S}\\
  M, \neighbor{S}, s \models \dpriv{\epsilon}{\delta} \phi
  & \textmd{ if } &
  \textmd{for all } t \textmd{ with } s \neighbor{S} t \textmd{ and
   position-independent scheduler } \\
  & & \scheduler{S},
   \myPr{s}{M_{\scheduler{S}}}{\neighbor{S}}{\phi} \leq \epsilon \cdot
   \myPr{t}{M_{\scheduler{S}}}{\neighbor{S}}{\phi} + \delta
   \textmd{ and }
   \myPr{t}{M_{\scheduler{S}}}{\neighbor{S}}{\phi} \leq \epsilon \cdot
  \\
  & &
   \myPr{s}{M_{\scheduler{S}}}{\neighbor{S}}{\phi} + \delta
\end{eqnarray*}

\lz{how about using "query sequence" instead of position-independent scheduler, and then argue that it makes sense for the resulting MDP model from differential privacy? This corresponds to a special scheduler that is applied on both neighboring states. }
Recall that $M_{\scheduler{S}}$ is but a Markov chain. The semantics
of $M_{\scheduler{S}}, \neighbor{S}, \pi \models \phi$ and hence the
probability $\myPr{s}{M_{\scheduler{S}}}{\neighbor{S}}{\phi}$ are
defined as in Markov chains.
The semantics of $\dpCTL$ on reactive Markov decision processes
is again standard except the differentially private operator
$\dpriv{\epsilon}{\delta}$. For any path formula $\phi$,
$\dpriv{\epsilon}{\delta} \phi$ specifies states which are $(\epsilon,
\delta)$-close to all its neighbors in having paths satisfying $\phi$
for arbitrary position-independent schedulers. That is, no
position-independent scheduler can force any of neighbors to have
probabilistically distinguishable behaviors.

\noindent
\emph{Discussion.}
Another definition of the differentially private operator
might be:
\begin{eqnarray*}
  M, \neighbor{S}, s \models \dpriv{\epsilon}{\delta}^{\mathit{bad}} \phi
  & \textmd{ if } &
  \textmd{for all } t \textmd{ with } s \neighbor{S} t \textmd{ and
  scheduler } \scheduler{S},
  \myPr{s}{M_{\scheduler{S}}}{\neighbor{S}}{\phi} \leq
  \epsilon \cdot\\
  && \myPr{t}{M_{\scheduler{S}}}{\neighbor{S}}{\phi} + \delta \textmd{ and }
  \myPr{t}{M_{\scheduler{S}}}{\neighbor{S}}{\phi} \leq
  \epsilon \cdot \myPr{s}{M_{\scheduler{S}}}{\neighbor{S}}{\phi}
  + \delta
\end{eqnarray*}
A state satisfies $\dpriv{\epsilon}{\delta}^{\mathit{bad}} \phi$ if
no scheduler could differentiate the probabilities of paths satisfying
$\phi$ from neighbors. Such a definition would allow strategies to
take different actions from different states. It might need different
action sequences to distinguish neighbors. Two neighbors could still be
differentiated by a action sequence and lose their privacy. We hence
prefer the original definition.

\hide{
Since the scheduler attaining the maximally probable
behavior from a state may be different from the scheduler attaining
the maximally probable behavior from its neighbors, two
neighbors may still be distinguished by a scheduler. The
weaker definition does not preserve differential privacy. We hence
prefer the original definition.
}

>>>>

Other than the differentially private operator, the semantics of
$\dpCTLstar$ is standard~\cite{BK:08:PMC}.
To intuit the semantics of $\dpriv{\epsilon}{\delta} \phi$,
recall that  $\myPr{s}{K}{N}{\phi}$ is the probability of having
paths satisfying $\phi$ from $s$. Thus, a state $s$ satisfies
$\dpriv{\epsilon}{\delta} \phi$ if having paths satisfying $\phi$
from $s$ is $(\epsilon, \delta)$-close to having paths satisfying
$\phi$ from every neighbor of $s$. In other words, it is
probabilistically similar to observe paths satisfying $\phi$ from $s$
and from all neighbors of $s$.

\section{Model Checking Algorithms}
We first describe the model checking algorithm for $\dpCTL$, which follows the classical
algorithms for $\PCTL$ by computing the states satisfying
sub state-formulae inductively~\cite{BK:08:PMC}. It hence suffices to
consider the inductive step where the states satisfying the subformula
$\dpriv{\epsilon}{\delta} (\phi)$ is computed.

@@

### 2018-01-27 syntax.tex
structures and some examples

Consider
\begin{eqnarray*}
  \Phi & ::= & \top \ |\ p \ |\ \neg \Phi \ |\ \Phi \wedge \Phi \ |\ 
               \PJ{J} \phi \ |\ \dpriv{\epsilon}{\delta} \phi\\
  \phi & ::= & \X \Phi \ |\ \Phi \until \Phi
\end{eqnarray*}
A \emph{state} formula $\Phi$ is either $\top$, an atomic proposition
$p$, the negation of a sate formula, the conjunction of two state
formulae, a \emph{probabilistic} operator $\PJ{J}$ with $J$
an interval in $[0, 1]$ followed by a path formula, or a
\emph{differentially private} operator $\dpriv{\epsilon}{\delta}$
 with two non-negative real numbers $\epsilon$ and $\delta$ followed
 by a path formula. A 
\emph{path} formula $\phi$ is a \emph{next} operator ($\X$)
followed by a state formula, or an \emph{until} operator
($\until$) operator enclosed by two state formulae.

>>>>

Consider
\begin{eqnarray*}
  \Phi & ::= & \top \ |\ p \ |\ \neg \Phi \ |\ \Phi \wedge \Phi \ |\
               \PJ{J} \phi \ |\ \dpriv{\epsilon}{\delta} \phi\\
  \phi & ::= & \Phi \ |\ \neg\phi  \ |\  \phi\wedge\phi  \ |\  \X \phi \ |\ \phi \until \phi
\end{eqnarray*}
A \emph{state} formula $\Phi$ is either $\top$, an atomic proposition
$p$, the negation of a sate formula, the conjunction of two state
formulae, a \emph{probabilistic} operator $\PJ{J}$ with $J$
an interval in $[0, 1]$ followed by a path formula, or a
\emph{differentially private} operator $\dpriv{\epsilon}{\delta}$
 with two non-negative real numbers $\epsilon$ and $\delta$ followed
 by a path formula. A
\emph{path} formula $\phi$ is simply a LTL formula, with temporal operator \emph{next}  ($\X$)
and  \emph{until} operator
($\until$) operator enclosed by two path formulae.

As in the classical setting, we consider the sublogic $\dpCTL$ by allowing only path formulae of the form $\X\Phi$ and $\Phi\until\Phi$.
